Clusterization Properties of D-optimal Designs in Infinite Dimensional Bayesian Linear Inverse Problems

### 2023-04-14 latex/vanishing.tex
Moved proofs to main text

\begin{equation}\label{eq:eigenproblem}
  \sigma^{-2}\fwd \postcov \fwd^* \obs^* = \obs^* \Xi,
\end{equation}
with $\Xi$ diagonal. Equation \eqref{eq:eigenproblem} looks like an
eigenvalue problem for the self-adjoint operator $\sigma^{-2}\fwd
\postcov \fwd^*$, where rows of $\obs$, namely $\meas_j,j=1,\dots, m$,
are eigenvectors. However, $\postcov$ depends on $\obs$ and thus, we
refer to \eqref{eq:eigenproblem} as a \emph{nonlinear} eigenvalue
problem. The following Proposition and Lemma are proved in the
appendix:

\begin{restatable*}[Two applications of Woodbury Matrix Identity]{proposition}{woodbury}\label{prop:twice woodbury}
  Assume $\fwd \prcov \fwd^*$ is invertible. Then
  \begin{align*}
    \begin{split}
      \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* 
      %
      %
      = \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1},
    \end{split}
  \end{align*}  
\end{restatable*}

\begin{restatable*}[Simultaneous diagonizability]{lemma}{simdiag}\label{lemma:sim diag}
  Let $\hil$ separable Hilbert space, $C:\hil \to \hil$ self-adjoint
  and $\func_1,\dots,\func_m \in \hil$. Denote $\func^*$ the element
  $\func$ acting as a linear functional. If
  \begin{equation*}
   (C + \sum_{j=1}^m \func_j\func_j^*) \func_l = \xi_l \func_l,\ l = 1,\dots,m
  \end{equation*}
  then $C$ and $\sum_{j=1}^m \func_j \func_j^*$ are simultaneously
  diagonalizable.
\end{restatable*}

@@

Before we prove Theorem \ref{thm:char} we record Lemma
\ref{lemma:free}, whose proof is delegated to the appendix:
\begin{restatable*}[Unit norm decomposition]{lemma}{free}\label{lemma:free}
  Let $M \in \R^{k \times k}$ symmetric positive definite with $\ttr M
  = m$, $m \geq k$. We can find $\func_j \in \R^k,j=1,\dots,m$
  with $\|\func_j\|=1$ and $A = (\func_1,\dots,\func_m)$ such that
  $AA^t = M$.
\end{restatable*}

>>>>

\begin{equation}\label{eq:eigenproblem}
  \sigma^{-2}\fwd \postcov \fwd^* \obs^* = \obs^* \Xi,
\end{equation}
with $\Xi$ diagonal. Equation \eqref{eq:eigenproblem} looks like an
eigenvalue problem for the self-adjoint operator $\sigma^{-2}\fwd
\postcov \fwd^*$, where rows of $\obs$, namely $\meas_j,j=1,\dots, m$,
are eigenvectors. However, $\postcov$ depends on $\obs$ and thus, we
refer to \eqref{eq:eigenproblem} as a \emph{nonlinear} eigenvalue
problem.

\begin{proposition}\label{prop:twice woodbury}
  Assume $\fwd \prcov \fwd^*$ is invertible. Then
  \begin{align*}
    \begin{split}
      \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* 
      %
      %
      = \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1},
    \end{split}
  \end{align*}  
\end{proposition}

\begin{proof}
  The proof amounts to using Woodbury's matrix identity twice and a
  regularization trick. The standard proof for Woodbury's matrix
  identity works in infinite dimensions, as long as all terms are well
  defined. Unfortunately, $\obs^*\obs$ is not invertible, so we force
  it to be. Recall Woodbury's matrix identity:
  $$
  (A + UCV)^{-1} = A^{-1} - A^{-1}U(C^{-1} + VA^{-1}U)^{-1}VA^{-1}.
  $$

  Denote $A := \prcov^{-1}$, $U := \fwd^*$, $V := \fwd$, $C :=
  \sigma^{-2} (\obs^*\obs+\eps I)$ for some $\eps > 0$. Then:
  %% \begin{align*}
  %%   \begin{split}
  \begin{equation}\label{eq:first}
  \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* (\obs^* \obs +\eps I) \fwd )^{-1}\fwd^* = \fwd ( \prcov - \prcov \fwd^* ( \sigma^2(\obs^*\obs + \eps I)^{-1} + \fwd \prcov \fwd^* )^{-1} \fwd \prcov ) \fwd^* \\
      %
      %
      %% &= X - X(\sigma^2Y_{\eps}^{-1} + X)^{-1}X
  %%   \end{split}
  %% \end{align*}
  \end{equation}
   Now denote $X := \fwd\prcov \fwd^*$ and $Y_{\eps} := \obs^*\obs +
   \eps I$, and $Y_{\eps}$ is invertible. Then \eqref{eq:first}
   becomes:
   \begin{equation}\label{eq:second}
     \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* (\obs^* \obs +\eps I) \fwd )^{-1}\fwd^* = X - X(\sigma^2Y_{\eps}^{-1} + X)^{-1}X
   \end{equation}

   Note that $X + \sigma^2 Y_{\eps}^{-1}$ is invertible, as the sum of
   two positive definite operators. Now, taking $A := X^{-1}, C :=
   \sigma^2Y_{\eps}^{-1}, U := I$ and $V := I$ and using Wodbury's
   matrix identity in reverse order:
  \begin{align*}
    \begin{split}
      \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* (\obs^* \obs +\eps I) \fwd )^{-1}\fwd^* &= X - X(\sigma^2Y_{\eps}^{-1} + X)^{-1}X \\
      %
      %
      %
      &= (X^{-1} + \sigma^{-2}Y_{\eps})^{-1} \\
      %
      %
      %
      &= (\fwd \prcov \fwd^* + \sigma^{-2} (\obs^*\obs + \eps I))^{-1}.
    \end{split}
  \end{align*}

  We conclude that $\forall \eps > 0$
  \begin{align*}
    \begin{split}
      \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* (\obs^* \obs +\eps I) \fwd )^{-1}\fwd^* 
     &= (\fwd \prcov \fwd^* + \sigma^{-2} (\obs^*\obs + \eps I))^{-1}.
    \end{split}
  \end{align*}
  Letting $\eps \to 0$ completes the proof.
\end{proof}

\begin{lemma}[Simultaneous diagonizability]\label{lemma:sim diag}
  Let $\hil$ separable Hilbert space, $C:\hil \to \hil$ self-adjoint
  and $\func_1,\dots,\func_m \in \hil$. Denote $\func^*$ the element
  $\func$ acting as a linear functional. If
  \begin{equation*}
   (C + \sum_{j=1}^m \func_j\func_j^*) \func_l = \xi_l \func_l,\ l = 1,\dots,m
  \end{equation*}
  then $C$ and $\sum_{j=1}^m \func_j \func_j^*$ are simultaneously
  diagonalizable.
\end{lemma}

\begin{proof}
  First, enumerate the eigenvalues of $C + \sum_{j=1}^m
  \func_j\func_j^*$ as $\xi_1,\dots,\xi_\ell$. Denote the
  indices of the eigenvectors corresponding to $\xi_i$
  \begin{equation*}
    S_i := \{ 1 \leq k \leq m | (C + \sum_{j=1}^m \func_j\func_j^* )\func_k = \xi_i \func_k \}.
  \end{equation*}
  Define further
  \begin{equation*}
    A_i := \sum_{k \in S_i} \func_k \func_k^*,
  \end{equation*}
  which is self-adjoint. Two observations are in order. First,
  $\sum_{j=1}^m \func_j\func_j^* = \sum_{i=1}^\ell A_i$. Second, $A_i
  \func_k = 0$ if $k\not \in S_i$, since eigenvectors of different
  eigenvalue are orthogonal. For $k \in S_i$
  \begin{equation}\label{eq:on vi}
    \xi_i \func_k = (C + \sum_{j=1}^m \func_j \func_j^* ) \func_k = (C + A_i) \func_k.
  \end{equation}
  Let $V_i := span \{\func_k \}_{k\in S_i}$. Observe that $V_i$ is
  invariant under $A_i$, by definition, and under $C$, by \eqref{eq:on
    vi}. Using \eqref{eq:on vi} again, we conclude that on $V_i$, $A_i
  = \xi_iI - C$. This immediately implies $A_i$ and $C$ have the
  same eigenvectors on $V_i$. This holds for every $1 \leq i \leq
  \ell$ and we conclude that $C$ and $A$ have the same eigenvectors.
\end{proof}

@@

For the proof of Theorem \ref{thm:char} we require Lemma
\ref{lemma:free}:

\begin{lemma}[Unit norm decomposition]\label{lemma:free}
  Let $M \in \R^{k \times k}$ symmetric positive definite with $\ttr M
  = m$, $m \geq k$. We can find $\func_j \in \R^k,j=1,\dots,m$
  with $\|\func_j\|=1$ and $A = (\func_1,\dots,\func_m)$ such that
  $AA^t = M$.
\end{lemma}

\begin{proof}
  Let us diagonalize $M$, so that $M = U D U^t$ with $D =
  \diag(d_1,\dots,d_k)$ and $U \in \R^{k \times k }$ orthogonal. Let
  $S \in \R^{k \times m}$ with $S_{ii} = \sqrt{d_{i}}$ and zeros
  otherwise. Define $A:= U S V^t$, where $V \in \R^{m \times m}$ is
  orthogonal and will be further restricted later. Then $AA^t = U
  SV^tVS^t U^t = UDU^t$, so $AA^t$ has the required eigenvalues and
  eigenvectors by construction. If we can choose $V$ such that $A$
  also satisfies the unit norm constraints we are done. These
  constraints are, for $j=1,\dots,m$:
  \begin{equation}\label{eq:V constraints}
   1 = [A^tA]_{jj} = [V S^tS V^t]_{jj},
  \end{equation}
  and we can expect to do this since we assumed $\ttr D = m$.

  Define $C = S^tS - I \in \R^{m \times m}$. Note that $\ttr C = 0$ and
  $C$ is diagonal with non-zero entries $d_i-1,i=1,\dots,k$. It suffices
  to find $V$ orthogonal such that $V C V^t$ has zero diagonal. We
  construct such $V$ by sequentially inserting zeros in the diagonal
  and not destroying zeros we already introduced, starting from the
  last diagonal entry and moving to the first. Since $c_{mm} \neq 0$ ,
  let $p < m$ such that $c_{pp}c_{mm} < 0$ (such $p$ exists because
  the trace is zero) and let $\theta \in (0,\pi)$. Define a Givens
  rotation $R^{(m)} \in \R^{m \times m}$ by
  \begin{equation*}
    r^{(m)}_{ab} :=
    \begin{cases}
      1 & a = b \neq p \text{ or } a = b \neq m \\
      \cos \theta & a = b = p  \\
     -\sin \theta & a = p, b = m\\
      \cos \theta & a = b = m \\
      \sin \theta & a = m, b = p \\ 
      0 & o.w
    \end{cases}
  \end{equation*}
  Note that conjugating a matrix by $R^{(m)}$ changes only its $m$ and
  $p$ rows and columns. We want to choose $\theta$ such that
  \begin{equation}\label{eq:mm}
    0 = [R^{(m)} C (R^{(m)})^t]_{mm} = \cos^2 \theta c_{mm} + 2\cos \theta \sin
    \theta c_{mp} + \sin^2\theta c_{pp},
  \end{equation}
  and it suffices to choose $\theta$ such that
  \begin{equation*}
    c_{mm} \cot^2 \theta + 2 c_{mp} \cot \theta + c_{pp} = 0.
  \end{equation*}
  This quadratic in $\cot\theta$ has a real solution, since
  $c_{pp}c_{mm} < 0$ by assumption and we can find $\theta \in
  (0,\pi)$ such that \eqref{eq:mm} is satisfied. We continue to find
  $R^{(m-1)}$ that leaves row and column $m$ unchanged and
  continue introducing zeros to the diagonal. The assumption $\ttr D =
  m \Rightarrow \ttr C = 0$ guarantees we can do that. Taking $V:=
  R^{(1)} R^{(2)} \dots R^{(m-1)}R^{(m)}$ completes the proof.
\end{proof}
